Curvature approximation on surfaces -- A Discrete Exterior Calculus Approach

### 2014-12-09 article/main.tex
...

\section{Curvature approximations}
  \subsection{Curvature vector} \label{secCurvatureVector}
    The curvature vector \( \vec{H} \) is given by
    \begin{align}
      \vec{H} = 2 H \vec{\nu}
    \end{align}
    for the mean curvature \( H \) and normals \( \vec{\nu} \). 
    In this paper, the mean curvature is set as the arithmetic mean of the two principal curvatures and
    \( H \) should be all over positive for a sphere (i.e. \( H\equiv1 \) on \( \mathds{S}^{2} \)).
    \( H \) and \( \vec{H} \) are extrinsic curvature values.
    This means, that an two dimensional ``inhabitant'' of the surface can't determine this kind of curvature 
    in contrast to the Gaussian curvatures.
    With the inclusion map \( \vec{x}:= \R^{3}\supset M \hookrightarrow \R^{3} \), which is the position vector of \( M \) in \( \R^{3} \),
    the curvature vector can be calculated by (see \cite[Ch. 5]{chen}, \cite[Ch. 4.5]{flanders} or \cite{elliott})
    \begin{align} \label{eqCurvatureVector}
      -\Delta_{B}\vec{x} = \vec{H} \formpunkt
    \end{align}
    (Note that \( \vec{x} \) is an isometric immersion with the common induced metric from \( \R^{3} \).)
    The equation \eqref{eqCurvatureVector} is component-wise to read and can be discretized with 
    discrete Laplace-Beltrami operator in \eqref{eqDisLaplace}, 
    where \( f \) becomes the \mh{i}{th} coordinate functions \( x^{i} \) for
    \( i=0,1,2 \).
    Hence, we can easily build the local coefficients matrices for the elements \( \sigma^{2} \) 
    with \( A^{\sigma^{2}} \) from \eqref{eqLokalMatrixLaplace}:
    \begin{align}
      \begin{bmatrix}
        A^{\sigma^{2}} & 0 & 0 \\
        0 & A^{\sigma^{2}} & 0 \\
        0 & 0 & A^{\sigma^{2}}
      \end{bmatrix}
    \end{align}
    to determine the left-hand side of the system
    \begin{align} \label{eqDisCurvatureVector}
      \left\langle *\Delta_{B} x^{i}, \star v \right\rangle
        &= \left\langle *H^{i}, \star v \right\rangle \formpunkt
    \end{align}
    for all vertices \( v\in K \) and \( i=0,1,2 \).
    \( H^{i}\in \R \) are the components of the curvature vector.
    The right-hand side of \eqref{eqDisCurvatureVector} can be locally represented by the decomposed Voronoi area 
    \begin{align}
      C^{\sigma^{2}}_{i} = \left|  \right|
    \end{align}
    around \( v \).

  \subsection{Weingarten map} \label{secWeingarten}

@@

  \begin{figure}
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/sphere/L2K.eps}
    \end{minipage}\hfill
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/sphere/L2H.eps}
    \end{minipage}\\
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/sphere/LMaxK.eps}
    \end{minipage}\hfill
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/sphere/LMaxH.eps}
    \end{minipage}
    \caption{Sphere \eqref{eqSphere}: Relative \( L_{2} \) (top) resp. \( L_{\infty} \) (bottom) error for \( K \) (left) and
                                                     \( H \) (right) for different \( h \) in a log-log-plot.}
    \label{figSphere}
  \end{figure}

>>>>

\section{Curvature approximations}
  \subsection{Curvature vector} \label{secCurvatureVector}
    The curvature vector \( \vec{H} \) is given by
    \begin{align}\label{eqMeanCurvCurvVec}
      \vec{H} = 2 H \vec{\nu}
    \end{align}
    for the mean curvature \( H \) and normals \( \vec{\nu} \). 
    In this paper, the mean curvature is set as the arithmetic mean of the two principal curvatures and
    \( H \) should be all over positive for a sphere (i.e. \( H\equiv1 \) on \( \mathds{S}^{2} \)).
    \( H \) and \( \vec{H} \) are extrinsic curvature values.
    This means, that an two dimensional ``inhabitant'' of the surface can't determine this kind of curvature 
    in contrast to the Gaussian curvatures.
    With the inclusion map \( \vec{x}:= \R^{3}\supset M \hookrightarrow \R^{3} \), which is the position vector of \( M \) in \( \R^{3} \),
    the curvature vector can be calculated by (see \cite[Ch. 5]{chen}, \cite[Ch. 4.5]{flanders} or \cite{elliott})
    \begin{align} \label{eqCurvatureVector}
      -\Delta_{B}\vec{x} = \vec{H} \formpunkt
    \end{align}
    (Note that \( \vec{x} \) is an isometric immersion with the common induced metric from \( \R^{3} \).)

    The equation \eqref{eqCurvatureVector} is to read component-wise and can be discretized with 
    the discrete Laplace-Beltrami operator in \eqref{eqDisLaplace}, 
    where \( f \) becomes the \mh{i}{th} coordinate functions \( x^{i} \) for
    \( i=0,1,2 \).
    I.e. we solve for all vertices \( v\in K \) and \( i=0,1,2 \).
    \begin{align} \label{eqDisCurvatureVector}
      \left\langle *\Delta_{B} x^{i}, \star v \right\rangle
        &= \left\langle *H^{i}_{LX}, \star v \right\rangle \formpunkt
    \end{align}
    \( H^{i}_{LX}\in \R \) are the components of the approximated curvature vector.
    The right-hand side (RHS) of \eqref{eqDisCurvatureVector} can be locally represented by the decomposed Voronoi area 
    \begin{align}
      C^{\sigma^{2}}_{i} = \left| \star v \cap \sigma^{2}  \right|
    \end{align}
    around \( v\prec\sigma^{2} \).
    Hence, the local coefficients matrix of the RHS of \eqref{eqDisCurvatureVector}
    (and generally for all Zero-Order-Terms of scalar valued functions) becomes
    \begin{align}
      \begin{bmatrix}
        C^{\sigma^{2}}_{0} & 0 & 0 \\
        0 & C^{\sigma^{2}}_{1} & 0 \\
        0 & 0 & C^{\sigma^{2}}_{2}
      \end{bmatrix}
    \end{align}
    and can be assembly in a common way.

    The mean curvature \( H \) is determined with \eqref{eqMeanCurvCurvVec} by taking the euclidean \mh{\R^{3}}{norm}
    \begin{align}
      H_{LX} = \frac{s}{2}\left\| \vec{H}_{LX} \right\|_{\R^{3}}
    \end{align}
    on every vertex with respect to the sign \( s:K^{(0)}\rightarrow\left\{ +1,-1 \right\} \),
    which depends on the direction of \(  \vec{H}_{LX} \).
    If the approximated curvature vector point outward the polytop \( K \), then is \( s=+1 \) and vice versa.

  \subsection{Weingarten map} \label{secWeingarten}
    The Weingarten map (shape operator) \( S \)
    can be obtained by rising on time the indices of the second fundamental form \( \II \)
    (see \cite{FirstCourse}),
    i.e. \( S = (g)^{-1}\II \) with a metric tensor \( g \),
    or by the exterior derivatives of the normal components:
    \begin{align}
    \begin{aligned}\label{eqDefWeingarten}
      S: T_{\vec{x}}M &\rightarrow T_{\vec{x}}M \\
                    \vec{w} &\mapsto \exd\vec{\nu} \left( \vec{w} \right) \formpunkt
    \end{aligned}
    \end{align}
    In particularly, \( \exd\nu^{i} \left( \vec{w} \right) \) is the directional derivative of the \mh{i}{th} normal component \( \nu^{i} \) 
    along \( \vec{w} \).
    The eigenvalues of \( S \) are the principal curvatures \( \kappa_{1} \) and \( \kappa_{2} \).
    Note that the linear operator \( S \) by \eqref{eqDefWeingarten} as matrix have a column space in local coordinates 
    and depends on the choice of the parametrisation as a consequence.
    A more manageable formulation is the extended Weingarten map
    \begin{align}\label{eqDefExWeingarten}
      \bar{S}:= \nabla\vec{\nu} \formkomma
    \end{align}
    which can be defined on the entire tangential space of the \( \R^{3} \) (if the normal vectors of \( M \) are smoothly extended to them).
    The point is, that \( \bar{S} \), restricted to \( T_{\vec{x}}M \), is the Weingarten map \( S \) (see \cite[Pt. 2, Ch. 2]{kimura}),
    with the additive eigenvalue 0,
    and \eqref{eqDefExWeingarten} can be approximated by our discrete average gradient \eqref{eqGradientAverage},
    i.e. we calculate the \mh{i}{th} row \( \bar{s}_{i}\in\R^{3} \) of the discrete extended Weingarten map \( \bar{S}_{W}\in\R^{3\times 3} \) by
    \begin{align} \label{eqDisWeingarten}
      \left\langle *\overline{\nabla}\nu^{i} , \star v \right\rangle 
        &= \left\langle *\bar{s}_{i} , \star v \right\rangle
    \end{align}
    on all vertices \( v\in K \).
    For numerical reason we symmetrize \( \bar{S}_{W} \) by
    \begin{align}
      \bar{S}_{W}^{Sym} &:= \frac{1}{2}\left( \bar{S}_{W} + \bar{S}_{W}^{T} \right)
    \end{align}
    to get the same property like the smooth extended Weingarten map.
    To obtain the eigenvalues we use the QR-algorithm of the MTL4\footnote{Matrix Template Library 4}.
    The lowest eigenvalue regarding to the absolute value is omitted.
    So w.l.o.g.
    \begin{align}
      \left\{\kappa_{1},\kappa_{2}\right\} &\approx \left\{ \kappa_{W,1}, \kappa_{W,2} \right\} \\ 
                    &:= \left\{ \lambda\in\text{Eig}(\bar{S}_{W}^{Sym}) \middle|\ \exists\lambda_{0}\in\text{Eig}(\bar{S}_{W}^{Sym}): \left| \lambda \right| > \lambda_{0} \right\} 
    \end{align}
    Hence, the approximated Gauss and mean curvatures are
    \begin{align}
      K_{W} &:= \kappa_{W,1} \cdot \kappa_{W,2} \\
      H_{W} &:= \frac{1}{2}\left( \kappa_{W,1} + \kappa_{W,2} \right) \formpunkt
    \end{align}

    The equations \eqref{eqDisWeingarten} can be applied, only if we know the exact normals of the manifolds on the vertices, 
    e.g. if the surface \( M \) is described by the 0-level-set of a signed distance function \( \varphi:\R^{3}\rightarrow\R \),
    then is
    \begin{align}
      \vec{\nu} &= \frac{\nabla_{\R^{3}}\varphi}{\left\| \nabla_{\R^{3}}\varphi \right\|} \formpunkt
    \end{align}
    For arbitrary surface triangulations, where the smooth manifold is unknown, like in figure \ref{figBunnyHorseCurvature},
    we must approximate the normals on the vertices before.
    A simple way to do that, is to average the element normals \( \vec{\nu}^{\sigma^{2}} \) of a triangle \( \sigma^{2}\in K \)
    over the Voronoi area:
    \begin{align}
      \left\langle *\vec{\nu}_{\avn} , \star v \right\rangle
        &:= \sum_{\sigma^{2}\succ v} \left| \star v \cap \sigma^{2} \right| \vec{\nu}^{\sigma^{2}} \formpunkt
    \end{align}
    With the same procedure above, we get the approximated curvatures \( K_{W,\avn} \) and \( H_{W,\avn} \).

@@

  \begin{figure}
    \centering\includegraphics[width=.49\textwidth]{bilder/sphere/L2K.eps}
    \centering\includegraphics[width=.49\textwidth]{bilder/sphere/L2H.eps}
    \caption{Sphere \eqref{eqSphere}: Relative \( L_{2} \) (top) resp. \( L_{\infty} \) (bottom) error for \( K \) (left) and
                                                     \( H \) (right) for different \( h \) in a log-log-plot.}
    \label{figSphere}
  \end{figure}
